---
title: "CIS Data Processing"
output: github_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(sf)
knitr::opts_chunk$set(echo = TRUE, dpi = 300, fig.path = "README_")
```

This repo describes a pipeline for turning .zip files downloaded from the [CIS search tool](https://iceweb1.cis.ec.gc.ca/Archive/page1.xhtml) into usable data products that can be queried and loaded efficiently in R. Roughly,

- Add .zip files to zip/
- Run zip-collect.R
- Run gpkg-collect.R
- Run attrs-collect.R

Both zip-collect.R and gpkg-collect.R are sufficiently lazy such that they do not decompress/load anything that has already been converted. The result is a folder gpkg/ that contains easy-to-read (i.e., `sf::read_sf()`) .gpkg files and a highly queryable attribute table (read using `arrow::read_parquet()`).

Alternatively, you can use the provided attrs.parquet and/or zip/meta.csv to download a subset of the data (urls for each data source are provided).

```{r}
library(tidyverse)

read_csv("zip/meta.csv")
```

```{r}
sf::read_sf("gpkg/EA_1968-06-25.gpkg")
```

```{r}
arrow::read_parquet("attrs.parquet")
```

## Dataset details

A documentation of the SIGRID vector file format (i.e.,
.gpkg files in gpkg/) can be found
[here](https://www.jcomm.info/index.php?option=com_oe&task=viewDocumentRecord&docID=4439); [a guide specific to the Canadian Ice Service](https://www.canada.ca/en/environment-climate-change/services/ice-forecasts-observations/latest-conditions/archive-overview/information-about-data.html) is also available. See also the [Ice glossary](https://www.canada.ca/en/environment-climate-change/services/ice-forecasts-observations/latest-conditions/glossary.html) provided by the Canadian Ice Service. The column names are the same for files between 1968 and present, but in 2020 there was a shift in data formats and only some columns were retained. In 1983 the `R_*` columns stopped being used.

The `E_*` columns (probably for "egg") are present in all ice polygon files. These are string codes where the empty string `''` is used both for "blank" and "0".

- `E_CA` (Partial concentration of thickest ice): '', '1', '2', '3', '4', '5', '6', '7', '8', '9'
- `E_CB` (Partial concentration of second thickest ice): '', '1', '2', '3', '4', '5', '6', '7', '8', '9'
- `E_CC` (Partial concentration of the third thickest ice): '', '1', '2', '3', '4', '5', '6', '7', '8'
- `E_CD` (Stage of development of any remaining class of ice (corresponds to Sd): '', '1', '2', '3', '4', '5', '6'
- `E_CS`: '', '1', '10', '2', '3', '4', '5', '6', '7', '8', '9', '9+'
- `E_CT` (Total concentration): '', '0.', '1', '10', '2', '3', '4', '5', '6', '7', '8', '9', '9+'
- `E_FA` (Form of thickest ice): '', '1', '2', '3', '4', '5', '6', '7', '8', '9', 'X'
- `E_FB` (Form of second thickest ice): '', '1', '2', '3', '4', '5', '6', '7', '8', '9', 'X'
- `E_FC` (Form of third thickest ice): '', '0', '1', '2', '3', '4', '5', '6', '7', '8', 'X'
- `E_FD` (Form of any remaining class of ice): '' (blank everywhere)
- `E_FE`: '' (blank everywhere)
- `E_SA` Stage of development of thickest ice): '', '1', '1.', '3', '4', '4.', '5', '6', '7', '7.', '8.', '9.'
- `E_SB` (Stage of development of second thickest Ice): '', '1', '1.', '2', '4', '4.', '5', '6', '7', '7.', '8.', 'X'
- `E_SC` (Stage of development of third thickest Ice): '', '1', '1.', '2', '3', '4', '4.', '5', '7'
- `E_SD` (Stage of development of any remaining class of ice): '', '1', '4', '4.', '5', '7'
- `E_SE`: '' (blank everywhere)
- `E_SO` (stage of development of ice found in traces): '', '1.', '4', '4.', '5', '6', '7', '7.', '8.', '9.'

Numeric values (here '' is blank whereas 0.0 is 0):

- `N_CB` (numeric value for the concentration of icebergs): '', ' 0.2', '10.0'
- `N_CFY` (numeric value for the concentration of first year ice): '', ' 0.0', ' 0.3', ' 1.0', ' 1.3', ' 2.0', ' 2.3', ' 3.0', ' 3.3', ' 4.0', ' 4.3', ' 5.0', ' 5.3', ' 6.0', ' 6.3', ' 7.0', ' 7.3', ' 8.0', ' 8.3', ' 9.0', ' 9.3', ' 9.7', '10.0'
- `N_CG` (numeric value for the concentration of grey ice): '', ' 0.0', ' 0.3', ' 1.0', ' 2.0', ' 3.0', ' 4.0', ' 5.0', ' 6.0', ' 7.0', ' 8.0', ' 9.0', ' 9.7', '10.0'
- `N_CGW` (numeric value for the  concentration o grey white ice): '', ' 0.0', ' 0.3', ' 1.0', ' 2.0', ' 3.0', ' 4.0', ' 5.0', ' 6.0', ' 7.0', ' 8.0', ' 9.0', ' 9.7', '10.0'
- `N_CMY` (numeric value for the concentration of multi-year ice): '', ' 0.0', ' 0.3', ' 1.0', ' 2.0', ' 3.0', ' 4.0', ' 5.0', ' 6.0', ' 7.0', ' 8.0', ' 9.0', ' 9.7', '10.0'
- `N_CN` (numeric value for the concentration of new ice): '', ' 0.0', ' 0.3', ' 1.0', ' 2.0', ' 3.0', ' 4.0', ' 5.0', ' 6.0', ' 7.0', ' 8.0', ' 9.0', ' 9.7', '10.0'
- `N_COI` (numeric value for the concentration of old ice): '', ' 0.0', ' 0.3', ' 0.6', ' 1.0', ' 1.3', ' 2.0', ' 2.3', ' 3.0', ' 3.3', ' 4.0', ' 4.3', ' 5.0', ' 5.3', ' 6.0', ' 6.3', ' 7.0', ' 7.3', ' 8.0', ' 8.3', ' 9.0', ' 9.3', ' 9.7', '10.0'
- `N_CSY` (numeric value for the concentration of second year ice): '', ' 0.0', ' 0.3', ' 1.0', ' 2.0', ' 3.0', ' 4.0', ' 5.0', ' 6.0', ' 7.0', ' 8.0', ' 9.0', ' 9.7', '10.0'
- `N_CT` (numeric value for total concentration): '', ' 0.0', ' 0.2', ' 0.3', ' 1.0', ' 2.0', ' 3.0', ' 4.0', ' 5.0', ' 6.0', ' 7.0', ' 8.0', ' 9.0', ' 9.7', '10.0'
- `N_CYI` (numeric value for the concentration of young ice): '', ' 0.0', ' 0.3', ' 0.6', ' 1.0', ' 1.3', ' 2.0', ' 2.3', ' 3.0', ' 3.3', ' 4.0', ' 4.3', ' 5.0', ' 5.3', ' 6.0', ' 6.3', ' 7.0', ' 7.3', ' 8.0', ' 8.3', ' 9.0', ' 9.3', ' 9.7', '10.0'

Prior to 1983, the following "ratio codes" were used. These appear to correspond mostly to the "numeric" codes, but are no longer documented.

- `R_CFY`: '', '/', '0', '1', '2', '3', '4', '5', '6', '7', '8', '9', 'F'
- `R_CG`: '', '/', '0', '1', '2', '3', '4', '5', '6', '7', '8', '9', 'F'
- `R_CGW`: '', '/', '0', '1', '2', '3', '4', '5', '6', '7', '8', '9', 'F'
- `R_CMY`: '', '/', '0', '1', '2', '3', '4', '5', '6', '7', '8', '9', 'F'
- `R_CN`: '', '/', '0', '1', '2', '3', '4', '5', '6', '7', '8', '9', 'F'
- `R_CS`: '', ' 1', ' 2', ' 3', ' 4', ' 5', ' 6', ' 7', ' 8', ' 9', '10', '9+'
- `R_CSY`: '', '/', '0', '1', '2', '3', '4', '5', '6', '7', '8', '9', 'F'
- `R_CT`: '', ' 1', ' 2', ' 3', ' 4', ' 5', ' 6', ' 7', ' 8', ' 9', '10', '4', '9+'
- `R_PFY`: '', '/', '0', '1', '2', '3', '4', '5', '6', '7', '8', '9', 'F'
- `R_PG`: '', '/', '1', '2', '3', '4', '5', '6', '7', '8', '9', 'F'
- `R_PGW`: '', '/', '0', '1', '2', '3', '4', '5', '6', '7', '8', '9', 'F'
- `R_PMY`: '', '/', '0', '1', '2', '3', '4', '5', '6', '7', '8', '9', 'F'
- `R_PN`: '', '/', '0', '1', '2', '3', '4', '5', '6', '7', '8', '9', 'F'
- `R_PSY`: '', '/', '0', '1', '2', '3', '4', '5', '6', '7', '8', '9', 'F'
- `R_SFY`: '', '/', '0', '1', '2', '3', '4', '5', '6', '7', '8', '9', 'F'
- `R_SG`: '', '/', '0', '1', '2', '3', '4', '5', '6'
- `R_SGW`: '', '/', '0', '1', '2', '3', '5', '6'
- `R_SMY`: '', '/', '0', '1', '2', '3', '4', '5', '6', '7', '8', 'F'
- `R_SN`: '', '/', '0', '1', '2', '3', '4', '5', '6', '7', '8', '9', 'F'
- `R_SSY`: '', '/', '0', '1', '2', '3', '4', '6', 'F'

Except for the `R_*` codes (which end in 1983), the columns have been consistently used over time.

```{r col-coverage, fig.width=6, fig.height=12, warning=FALSE}
attrs <- arrow::read_parquet("attrs.parquet")

attr_sum_time <- attrs %>% 
  select(region, date, matches("^[A-Z]_[A-Z]{2,3}$")) %>% 
  group_by(region, date) %>% 
  summarise_all(~any(. != "")) %>%
  ungroup() %>% 
  pivot_longer(-c(region, date)) %>% 
  mutate(year = lubridate::year(date)) %>% 
  group_by(region, year, name) %>% 
  summarise(used = any(value))

attr_sum_time %>%
  filter(!is.na(used)) %>% 
  ggplot(aes(x = year, y = fct_rev(name), alpha = used)) +
  geom_raster() +
  facet_wrap(vars(region), ncol = 3) +
  theme_bw(9) +
  theme(legend.position = "bottom")
```



