---
title: "CIS Data Processing"
output: github_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(sf)
library(stars)
options(dplyr.summarise.inform = FALSE)
knitr::opts_chunk$set(echo = TRUE, dpi = 300, fig.path = "README_")
```

This repo is an archive of the entire ice polygon dataset from the Canadian Ice Service, which can be downloaded piecemeal using the [CIS search tool](https://iceweb1.cis.ec.gc.ca/Archive/page1.xhtml). Use zip/meta.csv search for the existence of an ice polygon map (the `url` column contains the URL from which it can be downloaded). Use file-download.R to download the whole data set (about 14.8 GB uncompressed, 4.8 GB zipped) and gpkg.collect.R to convert files to .gpkg layers (which load 50-100 times faster than the raw .e00 files).

All of the geometry files are .gitignored (too big for a GitHub repo), but (some) are available in the shared directory on Eric's server.

- zip/: Contains zip files downloaded from the search tool.
- file/: Where all zip files were extracted (2020-01-20 onward .zip files contain .zip files, which are also extracted to file/ as shapefiles). These can be downloaded using file-download.R.
- gpkg/: A raw conversion of .e00 and .shp files in file/ to geopackage format with human-readable names.
- gpkg-standardized/: Like gpkg/ but with numeric columns as numbers, blank values as missings (NULLs), identical column names for all files, and standardized to the CRS of the most recent CIS files (Lambert Conformal Conic, WGS84).
- tif/: Rasterized versions of gpkg-standardized/. Currently the only attribute considered is `N_CT` (total ice concentration as a numeric value). All .tif files have identical grid extent and size (per region). All .tif files are standardized to the CRS of the most recent CIS files (Lambert Conformal Conic, WGS84).

## Examples in R

Using the [tidyverse](https://tidyverse.org) (for working with tables; has [excellent learning materials](https://r4ds.had.co.nz)), [sf](https://r-spatial.github.io/sf) (for working with vector geometry), and [stars](https://r-spatial.github.io/stars) (for working with raster geometry with a time component).

```{r}
library(tidyverse)
library(sf)
library(stars)
```

Read a .gpkg file:

```{r}
(poly <- read_sf("gpkg-standardized/EA_1968-06-25.gpkg"))
poly %>% 
  filter(PNT_TYPE < 400) %>% 
  select(PNT_TYPE, N_CT) %>% 
  plot()
```

Extract values at points of interest (best to use vector data for this):

```{r}
gpkg_hb_files <- list.files("gpkg-standardized", "^HB", full.names = TRUE)
gpkg_hb <- gpkg_hb_files %>% 
  head(100) %>% 
  map_dfr(read_sf)

target_locations <- tibble(
  label = c("P1", "P2"),
  longitude = c(-56.47, -60.43),
  latitude = c(54.85, 57.13)
) %>% 
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) %>% 
  st_transform(st_crs(gpkg_hb))

(intersected <- st_intersection(target_locations, gpkg_hb))

ggplot(intersected, aes(date, N_CT, col = label)) +
  geom_line() +
  theme_bw()
```

Read a .tif file (note that `N_CT` values were multiplied by 10 to take advantage of storing them more efficiently as 8-bit integers):

```{r}
(rst <- read_stars("tif/n-ct/EA_1968-06-25_n-ct.tif"))
st_bbox(rst)
plot(rst)

# do math (can include other rasters)
plot(rst / 10)

# work with matrix version
mat <- as.matrix(rst)[[1]]
str(mat)
mean(mat, na.rm = TRUE)
```

Align to an existing data set using `st_warp()` (a wrapper around [GDAL warp](https://gdal.org/programs/gdalwarp.html)). See the resampling method section for resampling options.

```{r}
data("levitus", package ="ocedata")

sst <- st_as_stars(
  st_bbox(c(xmin = -180, ymin = -89.5, xmax = 180, ymax = 89.5), crs = 4326),
  nx = length(levitus$longitude), ny = length(levitus$latitude),
  # stars expects reversed in the latitude direction
  values = levitus$SST[, rev(seq_along(levitus$latitude))]
)

bbox_latlon <- st_bbox(
  c(xmin = -73, ymin = 60, xmax = -52, ymax = 80),
  crs = 4326
)

levitus_study_area <- st_crop(sst, bbox_latlon)
rst_study_area <- st_warp(rst, levitus_study_area, use_gdal = TRUE, method = "mode")

plot(levitus_study_area)
plot(rst_study_area)
```

## Dataset details

Without downloading any files, you can use zip/meta.csv, gpkg/meta.csv, gpkg-standardized/meta.csv, and attrs.parquet to examine file metadata.

```{r}
(zip_meta <- read_csv("zip/meta.csv", col_types = cols()))
```

On average, there is one file per region per week. In 1983 the coding for some columns was changed, in (roughly) 1997 the HB region was added, and in (roughly) 2004, the CIS started producing all maps on Monday with internally-consistent polygons between overlapping regions.

```{r date-coverage, fig.width=6, fig.height=12, warning=FALSE, echo=FALSE}
zip_meta %>%
  mutate(
    year = lubridate::year(date),
    week = lubridate::isoweek(date)
  ) %>% 
  group_by(region, year, week) %>% 
  summarise(n = n()) %>%
  ggplot(aes(x = week, y = year, fill = factor(n, levels = 1:5))) +
  geom_tile() +
  scale_fill_brewer(limits = factor(1:5)) +
  scale_y_reverse(breaks = seq(1960, 2030, by = 10)) +
  facet_grid(rows = vars(region), scales = "free_y", space = "free_y") +
  coord_cartesian(expand = FALSE) +
  theme_bw(8) +
  theme(legend.position = "bottom")
```

```{r}
(gpkg_standardized_meta <- read_csv(
  "gpkg-standardized/meta.csv", 
  col_types = cols()
))
```

Also included is the collected attribute table from all files as a parquet file (because it's too big as a compressed CSV). You can use this to query relevant files based on attribute values and/or an area of interest before downloading any files. The CRS for coordinates in attrs.parquet is the same as those in gpkg-standardized/ (Lambert Conformal Conic, WGS84).

```{r}
library(arrow)
(attrs <- read_parquet("attrs.parquet"))
```

### CRS

The files in the archive appear to specify at least 4 coordinate reference systems. In some cases a coordinate reference system was not specified. All coordinate systems are lambert conformal conics with standard parallels at 49 and 77 north latitude with a central meridian at 100 degrees west longitude (latitude of origin: 40 degrees north latitude), however, they differ with respect to the datum/ellipsoid used to define the meridians/standard parallels. Since 2020-01-20 (when the CIS started distributing shapefiles), the WGS84 datum has been used. Prior to this date, it is likely that NAD27 was used to define the lambert conformal conic (except for several weeks during 2014 in the Great Lakes region).

```{r}
gpkg_meta <- read_csv("gpkg/meta.csv", col_types = cols())
gpkg_meta %>%
  mutate(gpkg_crs = str_replace_all(gpkg_crs, "meters", "METERS")) %>% 
  count(gpkg_crs) %>% 
  rev()
```

### Attributes

A documentation of the SIGRID vector file format (i.e.,
.gpkg files in gpkg/) can be found
[here](https://www.jcomm.info/index.php?option=com_oe&task=viewDocumentRecord&docID=4439); [a guide specific to the Canadian Ice Service](https://www.canada.ca/en/environment-climate-change/services/ice-forecasts-observations/latest-conditions/archive-overview/information-about-data.html) is also available. See also the [Ice glossary](https://www.canada.ca/en/environment-climate-change/services/ice-forecasts-observations/latest-conditions/glossary.html) provided by the Canadian Ice Service. The column names are the same for files between 1968 and present, but in 2020 there was a shift in data formats and only some columns were retained. In 1983 the `R_*` columns stopped being used.

The `E_*` columns (probably for "egg") are present in all ice polygon files. These are string codes where the empty string `''` is used both for "blank" and "0".

- `E_CA` (Partial concentration of thickest ice): '', '1', '2', '3', '4', '5', '6', '7', '8', '9'
- `E_CB` (Partial concentration of second thickest ice): '', '1', '2', '3', '4', '5', '6', '7', '8', '9'
- `E_CC` (Partial concentration of the third thickest ice): '', '1', '2', '3', '4', '5', '6', '7', '8'
- `E_CD` (Stage of development of any remaining class of ice (corresponds to Sd): '', '1', '2', '3', '4', '5', '6'
- `E_CS`: '', '1', '10', '2', '3', '4', '5', '6', '7', '8', '9', '9+'
- `E_CT` (Total concentration): '', '0.', '1', '10', '2', '3', '4', '5', '6', '7', '8', '9', '9+'
- `E_FA` (Form of thickest ice): '', '1', '2', '3', '4', '5', '6', '7', '8', '9', 'X'
- `E_FB` (Form of second thickest ice): '', '1', '2', '3', '4', '5', '6', '7', '8', '9', 'X'
- `E_FC` (Form of third thickest ice): '', '0', '1', '2', '3', '4', '5', '6', '7', '8', 'X'
- `E_FD` (Form of any remaining class of ice): '' (blank everywhere)
- `E_FE`: '' (blank everywhere)
- `E_SA` Stage of development of thickest ice): '', '1', '1.', '3', '4', '4.', '5', '6', '7', '7.', '8.', '9.'
- `E_SB` (Stage of development of second thickest Ice): '', '1', '1.', '2', '4', '4.', '5', '6', '7', '7.', '8.', 'X'
- `E_SC` (Stage of development of third thickest Ice): '', '1', '1.', '2', '3', '4', '4.', '5', '7'
- `E_SD` (Stage of development of any remaining class of ice): '', '1', '4', '4.', '5', '7'
- `E_SE`: '' (blank everywhere)
- `E_SO` (stage of development of ice found in traces): '', '1.', '4', '4.', '5', '6', '7', '7.', '8.', '9.'

Numeric values (here '' is blank whereas 0.0 is 0):

- `N_CB` (numeric value for the concentration of icebergs): '', ' 0.2', '10.0'
- `N_CFY` (numeric value for the concentration of first year ice): '', ' 0.0', ' 0.3', ' 1.0', ' 1.3', ' 2.0', ' 2.3', ' 3.0', ' 3.3', ' 4.0', ' 4.3', ' 5.0', ' 5.3', ' 6.0', ' 6.3', ' 7.0', ' 7.3', ' 8.0', ' 8.3', ' 9.0', ' 9.3', ' 9.7', '10.0'
- `N_CG` (numeric value for the concentration of grey ice): '', ' 0.0', ' 0.3', ' 1.0', ' 2.0', ' 3.0', ' 4.0', ' 5.0', ' 6.0', ' 7.0', ' 8.0', ' 9.0', ' 9.7', '10.0'
- `N_CGW` (numeric value for the  concentration o grey white ice): '', ' 0.0', ' 0.3', ' 1.0', ' 2.0', ' 3.0', ' 4.0', ' 5.0', ' 6.0', ' 7.0', ' 8.0', ' 9.0', ' 9.7', '10.0'
- `N_CMY` (numeric value for the concentration of multi-year ice): '', ' 0.0', ' 0.3', ' 1.0', ' 2.0', ' 3.0', ' 4.0', ' 5.0', ' 6.0', ' 7.0', ' 8.0', ' 9.0', ' 9.7', '10.0'
- `N_CN` (numeric value for the concentration of new ice): '', ' 0.0', ' 0.3', ' 1.0', ' 2.0', ' 3.0', ' 4.0', ' 5.0', ' 6.0', ' 7.0', ' 8.0', ' 9.0', ' 9.7', '10.0'
- `N_COI` (numeric value for the concentration of old ice): '', ' 0.0', ' 0.3', ' 0.6', ' 1.0', ' 1.3', ' 2.0', ' 2.3', ' 3.0', ' 3.3', ' 4.0', ' 4.3', ' 5.0', ' 5.3', ' 6.0', ' 6.3', ' 7.0', ' 7.3', ' 8.0', ' 8.3', ' 9.0', ' 9.3', ' 9.7', '10.0'
- `N_CSY` (numeric value for the concentration of second year ice): '', ' 0.0', ' 0.3', ' 1.0', ' 2.0', ' 3.0', ' 4.0', ' 5.0', ' 6.0', ' 7.0', ' 8.0', ' 9.0', ' 9.7', '10.0'
- `N_CT` (numeric value for total concentration): '', ' 0.0', ' 0.2', ' 0.3', ' 1.0', ' 2.0', ' 3.0', ' 4.0', ' 5.0', ' 6.0', ' 7.0', ' 8.0', ' 9.0', ' 9.7', '10.0'
- `N_CYI` (numeric value for the concentration of young ice): '', ' 0.0', ' 0.3', ' 0.6', ' 1.0', ' 1.3', ' 2.0', ' 2.3', ' 3.0', ' 3.3', ' 4.0', ' 4.3', ' 5.0', ' 5.3', ' 6.0', ' 6.3', ' 7.0', ' 7.3', ' 8.0', ' 8.3', ' 9.0', ' 9.3', ' 9.7', '10.0'

Prior to 1983, the following "ratio codes" were used. These appear to correspond mostly to the "numeric" codes, but are no longer documented.

- `R_CFY`: '', '/', '0', '1', '2', '3', '4', '5', '6', '7', '8', '9', 'F'
- `R_CG`: '', '/', '0', '1', '2', '3', '4', '5', '6', '7', '8', '9', 'F'
- `R_CGW`: '', '/', '0', '1', '2', '3', '4', '5', '6', '7', '8', '9', 'F'
- `R_CMY`: '', '/', '0', '1', '2', '3', '4', '5', '6', '7', '8', '9', 'F'
- `R_CN`: '', '/', '0', '1', '2', '3', '4', '5', '6', '7', '8', '9', 'F'
- `R_CS`: '', ' 1', ' 2', ' 3', ' 4', ' 5', ' 6', ' 7', ' 8', ' 9', '10', '9+'
- `R_CSY`: '', '/', '0', '1', '2', '3', '4', '5', '6', '7', '8', '9', 'F'
- `R_CT`: '', ' 1', ' 2', ' 3', ' 4', ' 5', ' 6', ' 7', ' 8', ' 9', '10', '4', '9+'
- `R_PFY`: '', '/', '0', '1', '2', '3', '4', '5', '6', '7', '8', '9', 'F'
- `R_PG`: '', '/', '1', '2', '3', '4', '5', '6', '7', '8', '9', 'F'
- `R_PGW`: '', '/', '0', '1', '2', '3', '4', '5', '6', '7', '8', '9', 'F'
- `R_PMY`: '', '/', '0', '1', '2', '3', '4', '5', '6', '7', '8', '9', 'F'
- `R_PN`: '', '/', '0', '1', '2', '3', '4', '5', '6', '7', '8', '9', 'F'
- `R_PSY`: '', '/', '0', '1', '2', '3', '4', '5', '6', '7', '8', '9', 'F'
- `R_SFY`: '', '/', '0', '1', '2', '3', '4', '5', '6', '7', '8', '9', 'F'
- `R_SG`: '', '/', '0', '1', '2', '3', '4', '5', '6'
- `R_SGW`: '', '/', '0', '1', '2', '3', '5', '6'
- `R_SMY`: '', '/', '0', '1', '2', '3', '4', '5', '6', '7', '8', 'F'
- `R_SN`: '', '/', '0', '1', '2', '3', '4', '5', '6', '7', '8', '9', 'F'
- `R_SSY`: '', '/', '0', '1', '2', '3', '4', '6', 'F'

Except for the `R_*` codes (which end in 1983), the columns have been consistently used over time.

```{r col-coverage, fig.width=6, fig.height=12, warning=FALSE, echo=FALSE}
attr_sum_time <- attrs %>% 
  select(region, date, matches("^[A-Z]_[A-Z]{2,3}$")) %>% 
  group_by(region, date) %>% 
  summarise_all(~any(. != "")) %>%
  ungroup() %>% 
  pivot_longer(-c(region, date)) %>% 
  mutate(year = lubridate::year(date)) %>% 
  group_by(region, year, name) %>% 
  summarise(used = any(value))

attr_sum_time %>%
  filter(!is.na(used)) %>% 
  ggplot(aes(x = year, y = fct_rev(name), alpha = used)) +
  geom_raster() +
  facet_wrap(vars(region), ncol = 3) +
  theme_bw(9) +
  theme(legend.position = "bottom")
```
